`treats`: a modular `R` package for simulating trees and traits.

### 2024-02-18 inst/treats_paper.Rmd
final version of the paper

To answer our question, we also want to simulate an extinction event.
To do so, we can create two different `"events"` object with the function `make.events`.
The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
The second one will simulate a random extinction but based on trait values: after reaching time 15, all the species with positive trait values will go extinct.
Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e. half the species are expected to evolve a trait value above 0).
For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex.
Because of this stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic (Fig. 5).

```{r, message = FALSE}
## Creating a random mass extinction
random_extinction <- make.events(
    target       = "taxa",
    condition    = age.condition(15),
    modification = random.extinction(0.75))
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target = "taxa",
    condition = age.condition(15),
    modification = trait.extinction(x = 0, condition = `>=`))

set.seed(1)
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a selective extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = positive_extinction,
                   null.error = 100,
                   replicates = 50)
```
```{r simtreetraitsevents, echo = FALSE, fig.cap="Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct. The red line marks the extinction event.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Visualising the difference between both scenarios
## Random extinction
plot(drop.singles(sim_rand_extinction[[31]]), main = "Random extinction")
abline(v = 15, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)

## Selective extinction
plot(drop.singles(sim_trait_extinction[[41]]), main = "Positive extinction")
abline(v = 15, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
```

@@

>>>>

To answer our question, we also want to simulate an extinction event.
To do so, we can create two different `"events"` object with the function `make.events`.
The first one will simulate a random extinction after reaching 66 time units and then making three quarter (0.75) of the taxa go extinct.
The second one will simulate a random extinction but based on trait values: after reaching time 66, all the species with positive trait values will go extinct.
Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e. half the species are expected to evolve a trait value above 0).
For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex.
Because of this stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic (Fig. 5).

```{r, message = FALSE}
## Creating a random mass extinction
random_extinction <- make.events(
    target       = "taxa",
    condition    = age.condition(140-66),
    modification = random.extinction(0.75))
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target = "taxa",
    condition = age.condition(140-66),
    modification = trait.extinction(x = 0, condition = `>=`))

set.seed(123)
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a selective extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = positive_extinction,
                   null.error = 100,
                   replicates = 50)
```
```{r simtreetraitsevents, echo = FALSE, fig.cap="Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct. The red line marks the extinction event.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Visualising the difference between both scenarios
## Random extinction
plot(sim_rand_extinction[[42]], main = "Random extinction")
abline(v = 66, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
## Selective extinction
plot(sim_trait_extinction[[50]], main = "Positive extinction")
abline(v = 66, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
```

@@

## Conflict of interest statement
I declare no conflicts of interest.

### 2024-02-18 paper/treats_paper.Rmd
final version of the paper

## Repeatability and reproducibility
This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/inst/paper.Rmd).
The data used for the example above [@beck2014ancient] is available from the `dispRity` package [@dispRity].

>>>>

## Repeatability and reproducibility
This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/inst/treats_paper.Rmd).
The data used for the example above [@beck2014ancient] is available from the `dispRity` package [@dispRity].
